---
title: "Report from 2025 fieldwork data"
format: html
editor: visual
---
#Fieldwork

## Sediment sampling

We sampled 70 lakes collecting two sediment cores from each labelled A and B. From each sediment core two slices of sediments were collected from depths 0-1 cm and 1-2 cm. From the core A, we collected samples for DNA analyses. These samples were collected from 0-1 cm and have volume of 5 cc. Additionally, for each DNA we took one blank.  

Samples from core B 0-1 cm are in Fredericton. They will be used solely for chironomid analyses. Total volume is ~30 cc. Perhaps we will have leftovers from this.

Samples from core A 0-1, 1-2, core B 1-2 cm, and DNA samples are in Poznań (DNA samples are kept frozen, and the rest in fridge). From this we have already done:\
- LOI550 and 950 (core A 0-1cm)  
- sand grain counting (core A 0-1cm)  

Subsamples taken already for  analyses in the future according to the label on bags:  
From core A 0-1 cm
- Chiro1 (2 cc)  
- Chiro2 (2 cc)  
- pollen (1 cc) 
- cladocera (1 cc)  
- diatoms (1 cc)  
- CNS  (1 cc)  
- C and O isotopes (1 cc) 
- N and C isotopoec (1 cc)

From core A 1-2 cm  
- pollen sulfur isotopes (20 cc)  

Samples Chiro1 and Chiro2 combined with leftover from core A samples 0-1 cm (very little material left) can be uses for preparation of XRF samples (Chiro1 and Chiro2 were collected before samples from core B 0-1 cm were sent to Fredericton for chironomid analyses).  

Apart from this we still have a lot of material from core B 1-2 cm as backup

## Water sampling

At coring sites, we took measurements with YSI instrument of water parameters. Two instruments were used; first was from Fredericton, second was from Poznań. For detail explanation of what was measured with what consult "fieldwaork_database.xlsx" file.

Along with this, water samples were taken approximately at the depth of several tens of cm. These were all kept cold expcept perhaps of shipping time (approx. 2-3 days). The analyses were done in RCP in Fredericton. 


TUTAJ DALEJ ZROBIC, NAJPIERW OPISAC, CO OPISYWALISMY W TERENIE, POTEM SAMA NALIZA DANYCH POPRZEDZONA MAPA.

## 

When you click the **Render** button a document will be generated that includes both content and the output of embedded code. You can embed code like this:

```{r}
1 + 1
```

You can add options to executable code like this

```{r}
#| echo: false
2 * 2
```

The `echo: false` option disables the printing of code (only output is displayed).
